---
title: "Moving Beyond Linearity"
output: github_document
---

## Background

Linear models are relatively simple to describe and implement, and have advantages over other approaches in terms of interpretation and inference. However, standard linear regression can have limitations in terms of predictive power because the assumption of linearity is almost always an approximation, and sometimes a poor one. Linearity (the relationship between the predictors and the outcome is linear) is one of the assumptions of linear regression model. There are several statistical methods to handle the non linearity. Most sophisticated is the extension of linear model like polynomial regression, step functions and splines.

## Polynomial Regression

**Polynomial regression** extends the linear model by adding extra predictors, obtained by raising each of the original predictors to a power. For example, a cubic regression uses three variables, $x$, $x^2$, and $x^3$, as predictors. This approach provides a simple way to provide a non-linear fit to data.

The relationship between the predictors and the response is non-linear has been to replace the standard linear model

$$
y_i = β_0 + β_1x_i + ϵ_i
$$

with a polynomial function

$$
y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^3 + \beta_3x_i^3 + ... + \beta_dx_i^d + \epsilon_i
$$

where ϵi is the error term.

For large enough degree d, a polynomial regression allows us to produce an extremely non-linear curve. Generally, it is unusual to use d greater than 3 or 4 because for large values of d, the polynomial curve can become overly flexible and can take on some very strange shapes.

# Practical Example

We will be analysing complex data ***Wage***. We begin by loading the ISLR2 library, which contains the data. This data is has 3000 observation of 11 different variables which includes year, age, marital status, race, logwage, wage, among other.

```{r, warning=F}
library(ISLR2)
library(ggplot2)
wage <- Wage
```

# Assessing non-linearity by visualization

```{r, warning=F}
ggplot(data = wage, aes(x = age, y = wage)) + 
  geom_point()
 
# Fitting linear regression

ggplot(data = wage, aes(x = age, y = wage)) + 
  geom_point() + geom_smooth(method = "lm") +
  labs(title = "Linear Regression of Wage in Age")
```

# Fitting polynomial regression model

```{r, warning=F}
polyfit <- lm(wage ~ poly(age, 4), data = wage)
coef(summary(polyfit))
```

This syntax fits a linear model, using the $lm()$ function, in order to predict wage using a fourth-degree polynomial in age: $poly(age, 4)$.

\###**Visualizing our fitted model**

```{r, warning=F}

# Create a new data frame for prediction
age_grid <- data.frame(age = seq(min(wage$age), max(wage$age), length.out = 100))

# Predict fitted values
age_grid$predicted_wage <- predict(polyfit, newdata = age_grid)

# Plot
ggplot(wage, aes(x = age, y = wage)) +
  geom_point(alpha = 0.5) +                     # original points
  geom_line(data = age_grid, aes(x = age, y = predicted_wage), color = "blue", size = 1.2) +
  labs(title = "4th-degree Polynomial Regression of Wage on Age",
       x = "Age", y = "Wage") +
  theme_minimal()

```

### **How to decide the degree of polynomial to use?**

In performing a polynomial regression we must decide on the degree of the polynomial to use. You can do Cross validation, Likelihood Ratio test, hypothesis test, check AIC. One way to do this is by using hypothesis tests. We now fit models ranging from linear to a degree-5 polynomial and seek to determine the simplest model which is suﬀicient to explain the relationship between wage and age. We use the $anova()$ function, which performs an analysis of variance (ANOVA, using an F-test) in order to test the null hypothesis that a model M1 is suﬀicient to explain the data against the alternative hypothesis that a more complex model M2 is required.

```{r}

poly.M1 <- lm(wage ~ age, data = wage)
poly.M2 <- lm(wage ~ poly(age, 2), data = wage)
poly.M3 <- lm(wage ~ poly(age, 3), data = wage)
poly.M4 <- lm(wage ~ poly(age, 4), data = wage)
poly.M5 <- lm(wage ~ poly(age, 5), data = wage)

anova(poly.M1,poly.M2,poly.M3,poly.M4,poly.M5)


```

The ANOVA comparison indicates that the polynomial terms upto degree 3 significantly improve model fit. Higher order terms beyond 3 do not provide significant meaningful results, suggesting that the cubic polynomial is sufficient.

## Step Function

**Step functions** cut the range of a variable into K distinct regions in order to produce a qualitative variable. This has the effect of fitting a piece wise constant function. We break the range of X into bins, and fit a different **constant** in each bin.

## Regression Splines

**Regression splines** are more flexible than polynomials and step functions, and in fact are an extension of the two. They involve dividing the range of $X$ into K distinct regions. Within each region, a polynomial function is fit to the data. However, these polynomials are constrained so that they join smoothly at the region boundaries, or knots. Provided that the interval is divided into enough regions, this can produce an extremely flexible fit.

Divide $X$ into smaller sub-intervals,

$$
x_{min} = t_0 < t_1 < .... t_m < t_{m+1} = x_{max}
$$

The breakpoints $t_1$, $t_m$, $t_{m +1}$ are called knots.

We construct $m + 1$ intervals, and for each sub intervals, we apply $m + 1$ polynomial functions as mentioned below:

$$ 
f_i(x) = β_{i,0} + β_{i,1}x_1 + β_{i,2}x_2 + β_{i,3}x_3 ....
$$

for $$ i \in [t_0, t_1], [t_1, t_2], ... [t_m, t_{m+1}] $$

where, $x_1, x_2, x_3$ are corresponding polynomial covariates.

## Constraint

**In the next step**, we must ensure that the fitted functions should be continuous and smooth at each knot. In other words, there should not be sudden jump or sudden fall in our function at each interior knot. To solve this, if we have K knots, this is done by making continuous derivative up to (K-1) order. Assumption of continuous derivatives ensures that polynomial pieces are joining together smoothly at all interior knot points.

A polynomial spline of degree p is a function that

-   Consists of a polynomial of degree p within each of the intervals, and

-   Has continuous derivatives up to the order (p−1) at each knot.

-   e.g., a cubic spline consists of polynomial transformations of degree 3 within each interval and has 2 times continuous derivatives.

## Regression splines using basis functions

Regression splines may initially seem complicated because they involve fitting piecewise polynomials while ensuring that the function (and some of its derivatives) remain continuous at the knots. Fortunately, this complexity can be handled using a basis function representation.

A cubic regression spline with K knots can be written as a linear regression model:

$$
y_i = \beta_0 + \beta_1 b_1(x_i) + \beta_2 b_2(x_i) + \cdots + \beta_{K+3} b_{K+3}(x_i) + \varepsilon_i
$$

where:

-   $b_1(x), \dots, b_{K+3}(x)$ are spline basis function

-   $\beta_0, \dots, \beta_{K+3}$ are unknown coefficients

-   $\varepsilon_i$ is the error term

## **Truncated power basis for cubic splines**

To construct a cubic spline:

-   Start with a cubic polynomial basis: x, x\^2, and x\^3

-   Add one truncated power basis function for each knot

A truncated power basis function is defined as:

$$
h(x, \xi) = (x - \xi)_+^3 =
\begin{cases}
(x - \xi)^3, & x > \xi \\
0, & x \le \xi
\end{cases}
$$

In order to fit a cubic spline to a data set with K knots, we perform least squares regression with an intercept and 3 + K predictors, of the form $X, X^2, X^3,h(x,\xi_1)$, $h(x, \xi_2), ... h(x, \xi_K)$, where $ξ_1, . . . , ξ_K$ are the knots.

## **Restricted/Natural Splines**

Regression splines can exhibit high variance near the boundaries of the predictor space, that is, when X takes very small or very large values. In these regions, the fitted curve may behave erratically, leading to wide and unstable confidence intervals.

A natural spline is a regression spline with additional boundary constraints: function is required to be linear at the boundary (in the region where X is smaller than the smallest knot, or larger than the largest knot). This additional constraint means that natural splines generally produce more stable estimates at the boundaries.


























